\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Smart Placement Portal: An Intelligent Recruitment Management System\\
}

\author{\IEEEauthorblockN{User Name}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University Name}\\
City, Country \\
email@example.com}
}

\maketitle

\begin{abstract}
The recruitment process in educational institutions involves managing a vast amount of data regarding students, companies, and job applications. Manual handling of this data is error-prone and inefficient. This paper presents the ``Smart Placement Portal,'' a comprehensive web-based application developed using the MERN stack (MongoDB, Express.js, React.js, Node.js). The system streamlines the entire placement workflow, offering distinct interfaces for Students, Companies, and Administrators. Key features include secure authentication with OTP verification, dynamic job posting with deadline enforcement, real-time application tracking, and an intelligent recommendation system that matches students to jobs based on their skills and resume data. The proposed system ensures transparency, efficiency, and an enhanced user experience through a modern, responsive user interface.
\end{abstract}

\begin{IEEEkeywords}
Web Application, MERN Stack, Recruitment, Machine Learning, Automation, Placement Portal.
\end{IEEEkeywords}

\section{Introduction}
\IEEEPARstart{P}{lacement} activities are crucial for academic institutions, bridging the gap between students and the industry. Traditional methods often utilize spreadsheets or disjointed systems, leading to communication gaps and data redundancy. The Smart Placement Portal addresses these challenges by centrally managing the entire lifecycle of the recruitment process.

The primary objective is to facilitate seamless interaction between students seeking opportunities and companies looking for talent. By integrating a recommendation engine, the system goes beyond simple data management to assist students in finding the most relevant opportunities, thereby increasing placement success rates.

\section{System Architecture}
The application follows a standard Client-Server architecture, utilizing the MERN stack for full-stack development.

\subsection{Frontend (Client)}
The user interface is built using \textbf{React.js}, ensuring a dynamic Single Page Application (SPA) experience.
\begin{itemize}
    \item \textbf{Design}: A premium ``Light Theme'' design system is implemented using CSS3 variables, featuring glassmorphism, smooth animations, and a responsive grid layout.
    \item \textbf{State Management}: React Hooks (useState, useEffect, useContext) manage local and global application states.
    \item \textbf{Interaction}: Axios is used for handling HTTP requests to the backend API.
\end{itemize}

\subsection{Backend (Server)}
The server-side logic is powered by \textbf{Node.js} and \textbf{Express.js}.
\begin{itemize}
    \item \textbf{API}: RESTful API endpoints handle data operations (CRUD) for users, jobs, and applications.
    \item \textbf{Security}: Implementation of JSON Web Tokens (JWT) for stateless authentication, removing the need for server-side sessions.
    \item \textbf{Middleware}: Custom middleware handles Role-Based Access Control (RBAC) to secure routes (e.g., ensuring only companies can post jobs).
\end{itemize}

\subsection{Database}
\textbf{MongoDB} is used as the NoSQL database. It stores unstructured data flexibly, which is ideal for varying job descriptions and user profiles. Mongoose ODM is used for schema validation (e.g., strictly enforcing job deadlines).

\section{Key Modules and Features}

\subsection{Authentication Module}
Security is paramount. The system implements:
\begin{itemize}
    \item \textbf{Registration}: Captures user details and role (Student/Company).
    \item \textbf{OTP Verification}: Ensures email validity before account activation.
    \item \textbf{Password Encryption}: Uses bcryptjs to hash passwords before storage.
\end{itemize}

\subsection{Student Module}
Designed to empower students in their job hunt.
\begin{itemize}
    \item \textbf{Dashboard}: Displays Profile Score, Application Stats, and Recommendations.
    \item \textbf{Resume Management}: Allows students to upload, preview, and update their resumes.
    \item \textbf{Job Application}: Students can view details and apply for jobs. Once applied, the job moves from the ``All Opportunities'' list to the ``My Applications'' list to prevent duplication.
    \item \textbf{Status Tracking}: Real-time updates on application status (Pending, Accepted, Rejected) are visible immediately.
\end{itemize}

\subsection{Company Module}
Provides tools for recruiters to manage their hiring drive.
\begin{itemize}
    \item \textbf{Job Posting}: Companies can create job listings. The system enforces a mandatory deadline field to ensure validity.
    \item \textbf{Applicant Management}: Recruiters can view applicants for their jobs, download resumes, and update application statuses.
    \item \textbf{Data Privacy}: Companies can only view applications for their own posted jobs.
\end{itemize}

\subsection{Recommendation Engine}
An intelligent subsystem matches students to jobs.
\begin{itemize}
    \item \textbf{Logic}: It compares the skills extracted from the student's profile/resume against the \texttt{skillsRequired} field of active job listings.
    \item \textbf{Filtering}: The engine automatically filters out jobs that have already been applied to or whose deadline has passed.
\end{itemize}

\section{Implementation Details}

\subsection{Deadline Enforcement}
To maintain data integrity, the system enforces job deadlines at two levels:
\begin{enumerate}
    \item \textbf{Backend Controller}: Rejects any job creation request missing a deadline.
    \item \textbf{Database Schema}: The Mongoose schema marks the deadline field as \texttt{required}, preventing invalid writes at the database level.
\end{enumerate}

\subsection{Dynamic Theme System}
The application features a modern UI system based on CSS variables:
\begin{verbatim}
:root {
  --primary: #2563eb; /* Royal Blue */
  --bg-gradient: linear-gradient(135deg,
                 #f0f9ff 0%, #e0f2fe 100%);
}
\end{verbatim}
This allows for consistent styling and easy maintenance.

\section{Conclusion}
The Smart Placement Portal successfully automates the campus recruitment process. By integrating secure authentication, robust data management, and intelligent features like recommendations and real-time status tracking, the system significantly reduces manual effort for placement officers and recruiters while providing a seamless experience for students. Future enhancements may include automated interview scheduling and advanced analytics for placement trends.

\begin{thebibliography}{00}
\bibitem{b1} MongoDB Inc., ``MongoDB Documentation,'' https://docs.mongodb.com.
\bibitem{b2} Facebook, ``React Documentation,'' https://reactjs.org.
\bibitem{b3} OpenJS Foundation, ``Node.js Documentation,'' https://nodejs.org.
\bibitem{b4} REST API Design Rulebook, O'Reilly Media.
\end{thebibliography}

\end{document}
